---
title: "Step Analysis - Reproducible Research Project 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is an analysis of data acquired from a personal activity monitoring device. The data are from one  
individual and has the following variables:  

* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)  
* **date**: The date on which the measurement was taken in YYYY-MM-DD format  
* **interval**: Identifier for the 5-minute interval in which measurement was taken  

There are a total of 17,568 observations in this dataset.  

We will be using the following packages (we will suppress warnings to keep the html compact):
```{r pkgs, results="hide",warning=FALSE}
suppressWarnings(suppressMessages(library(plyr)))
suppressWarnings(suppressMessages(library(dplyr)))
suppressWarnings(suppressMessages(library(data.table)))
suppressWarnings(suppressMessages(library(downloader)))
suppressWarnings(suppressMessages(library(xtable)))
suppressWarnings(suppressMessages(library(lubridate)))
suppressWarnings(suppressMessages(library(ggplot2)))
```

## Loading and preprocessing the data

Next we will load the data into a dataframe, rename and summarize.
```{r loadtable}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, "Steps_Dataset.zip")
unzip("Steps_Dataset.zip")
activity <- read.table('activity.csv', sep = ",", header = TRUE)
head(activity)
```

## Mean total number of steps taken per day

First, let's check the total number of steps per day.
```{r meaner}
meancheck <- activity %>%
  group_by(date) %>%
  summarise(totaldaily = sum(steps, na.rm=TRUE))
```

Second, we will create a histogram of the total steps per day.
```{r meanhist}
meancheck$totaldaily <- as.numeric(meancheck$totaldaily)
hist(meancheck$totaldaily, breaks=100, col="red")
```


Finally, let's calculate the mean and median
```{r meandian}
mean_steps <- meancheck %>%
  summarise(mean_total = mean(totaldaily, na.rm=TRUE))
mean_steps

median_steps <- meancheck %>%
  summarise(median_total = median(totaldaily, na.rm=TRUE))
median_steps
```

## Average daily activity pattern

Here we will group by interval but also take the average across days while doing so.
```{r meanplot}
interval_steps <- activity %>%
  group_by(interval) %>%
  summarise(totaldaily = mean(steps, na.rm=TRUE))
interval_steps

plot(interval_steps$interval, interval_steps$totaldaily, type = "l", 
     xlab="Interval", ylab="Average daily steps", col="red" , lwd=2)
```

And the maximum steps for an interval is:
```{r maxint}
max_interval <- interval_steps[which(interval_steps$totaldaily == max(interval_steps$totaldaily, na.rm = TRUE)), ]
max_interval
```

## Impute missing value

The total number of missing values is:
```{r miss}
nrow(activity[activity$steps == "NA",])
```

For imputation, We will use the average value for the interval across all days.
```{r imputer}
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
activity_imp <- ddply(activity, ~ interval, transform, steps = impute.mean(steps))
activity_impord <- activity_imp[order(activity_imp$date), ]
nrow(activity_imp[activity_impord$steps == "NA",])
```

And here is the new histogram as well as the new mean and median:
```{r imphist}
meancheck_imp <- activity_impord %>%
  group_by(date) %>%
  summarise(totaldaily = sum(steps, na.rm=TRUE))

meancheck_imp$totaldaily <- as.numeric(meancheck_imp$totaldaily)
hist(meancheck_imp$totaldaily, breaks=100, col="red")

mean_steps_imp <- meancheck_imp %>%
  summarise(mean_total = mean(totaldaily, na.rm=TRUE))
mean_steps_imp

median_steps_imp <- meancheck_imp %>%
  summarise(median_total = median(totaldaily, na.rm=TRUE))
median_steps_imp
```

## Difference between Weekdays and Weekends

First we need to add a variable that differentiates between weekends and weekdays.  
In this case, we will use a boolean flag.
```{r daytype}
activity_impord %>%
  mutate(weekend = (weekdays(ymd(date))=="Saturday" | weekdays(ymd(date))=="Sunday")) %>%
  head()
```

Next we will create two separate data sets for weekends and weekdays using this flag.  
First for the weekends:
```{r wknd}
int_wknd <- activity_impord %>%
  mutate(weekend = (weekdays(ymd(date))=="Saturday" | weekdays(ymd(date))=="Sunday")) %>%
  filter(weekend == TRUE) %>%
  group_by(interval) %>%
  summarise(totaldaily = mean(steps, na.rm=TRUE)) %>%
  print()
```
Then for the weekdays:
```{r wkdy}
int_wkdy <- activity_impord %>%
  mutate(weekend = (weekdays(ymd(date))=="Saturday" | weekdays(ymd(date))=="Sunday")) %>%
  filter(weekend == FALSE) %>%
  group_by(interval) %>%
  summarise(totaldaily = mean(steps, na.rm=TRUE)) %>%
  print()
```

Finally lets visually compare them by plotting the time trends of both.
```{r trendplot}
par(mfrow=c(2,1),mar=c(4,4,1,1)+0.1, cex = 0.75)
plot(int_wkdy$interval, int_wkdy$totaldaily, type = "l", 
     xlab="Interval", ylab="Weekday daily steps", col="red" , lwd=2)
plot(int_wknd$interval, int_wknd$totaldaily, type = "l", 
     xlab="Interval", ylab="Weekend daily steps", col="blue" , lwd=2)
```

This is the end of the analysis.